|

### 2024-07-24 manuscript/paper.Rmd
Yoel adding text to introduction, updating and formatting references, etc.

Ecological release theory suggests that a prey population’s niche should expand when predation is relaxed or removed, as prey take advantage of new enemy free space (reviewed in @Herrmann2021). Niche expansion could manifest through a combination of increasing within- and between-individual niche widths (@Bolnick2010; @Herrmann2021), including divergence between the sexes (@Bolnick2003; @Cooper2011). Theory and empirical data suggest that a population presented with ecological opportunity following ecological release might experience disruptive selection on males and females stemming from intraspecific competition over newly accessible resources, resulting in intersexual divergence in habitat use (@Schoener1968; @Shine1989; @Bolnick2003; @Butler2007; @Bolnick2008; @Cooper2011; but see @Stuart2021; @Blain2022). Accordingly, sexes should also diverge in morphological traits associated with habitat use; such ecologically-driven "character displacement between the sexes" is therefore one explanation for phenotypic sexual dimorphism (@deLisleRowe2015; @deLisleNco2018).

We tested this link between release from predators and the evolution of sexual dimorphism in a well-preserved, finely-resolved, ~16,000 year-long sequence of the fossil stickleback fish (<i>Gasterosteus doryssus</i>). The fossil lineage appeared in the depositional environment as a fully armored form, with complete pelvic girdles, two pelvic spines, and three dorsal spines, on average (@Bell2006; @Stuart2020). However, this habitat appears to have been missing fish piscivores (e.g., trout and other salmonids known to prey on modern stickleback; @Bell2009), relaxing putative selection for armor (@Reimchen1980, @Reimchen1994, @BellEtAl1993; @Bowne1994; @Roestietal2023). The lineage adaptively evolved armor loss but also reduction in the means of several other traits  (@Hunt2008; @Stuart2020; Siddiqui et al. in prep). In particular, tooth wear data suggest that the population shifted to eat more planktonic prey through time, indicative of increased open water habitat use (Purnell et al. 2007). If this divergence in habitat use is partitioned between sexes (@ReimchenNosil2004; @SpoljaricReimchen2008;@Reimchen2016), then we should expect sexual dimorphism to evolve for traits related to habitat use in <i>G. doryssus</i>.

To test this prediction, we imputed the sex of individual fossils, which is unobserved in practice, with Multiple Imputation by Chained Equations (MICE) (@MICE), using observed data from extant stickleback of known sex to train the imputation model.  Then we tracked the multivariate evolution of sexual dimorphism over ~16,000 years for 16 traits related to swimming, feeding and defense.

@@

## Fossil Specimen Data
We used Gasterosteus doryssus data that were previously reported by @Stuart2020, @Voje2022, and Siddiqui et al. (in review). Briefly, the data were collected from fossil Series K from Quarry D (Cerasoni et al., in review), dug from an open pit diatomite mine at 9.526° N, 119.094° W, near Fernley, Nevada, USA. Series K consisted of 18 samples taken at ~1000-year intervals, and mean sample times span ~16,363 years. Fish from series K were measured for 16 ecomorphological traits related to armor, swimming, and feeding (Table 1). Series K started at a previously documented horizon when a low armored lineage of stickleback with zero to one dorsal spines, zero pelvic spines, and highly reduced pelvises was completely replaced by a high armored lineage of stickleback with three dorsal spines, two pelvic spines, and complete (Bell et al. 2006; Bell 2009; Stuart et al. 2020). This lineage subsequently evolved reduction in armor, body size, and traits related to swimming and feeding (Bell et al. 2006; Stuart et al. 2020; Siddiqui et al. in review). The tempo and mode of armor reduction during this sequence suggests adaptive evolution by natural selection (Hunt et al. 2008), and we focus on the multivariate evolution of sexual dimorphism by this second lineage.

The fossil data consists of `r sum(is.na(stick$gender))` specimens with unknown sex over 18 time periods spaced  ~1000 years apart. Figure 1 shows the sample size at each of the 18 time periods.  There are at least 22 specimens at each time period with a high of 67 specimens in period 7.

```{r fig.cap = "Fossil sampel size for each time period"}
ggplot(aes(x = time, y = count), data = fortab) + geom_col() + xlab("Time Period")  + scale_x_continuous(breaks = 1:18, labels = 1:18)
```

@@

## Extant Specimen data
To span the gamut of stickleback diversity for our predictive model, we sampled modern stickleback from lakes containing generalist stickleback populations (@Hendry2009; @Bolnick2011) and from lakes containing benthic-limnetic species pairs (Baumgartner et al. 1988; @Schluter1992) (Table 2 (What is this?)). The generalist populations were collected by YES in 2013 and previously described in @Stuart2017. These samples were fixed in formalin, then stained for bone with Alizarin Red in 2013. Benthic and limnetic specimens were kindly loaned by D. Schluter and his lab at University of British Columbia. They collected benthic and limnetic individuals from Enos Lake in 1988 and from Emily Lake, Little Quarry Lake, Paxton Lake, and Priest Lake in 2018. The Enos specimens had been fixed whole in formalin and stored in 40% isopropanol. The specimens from the other lakes were initially preserved whole in 95% ethanol in the field before being gradually transferred to water then formalin in the lab and ultimately stored in 40% isopropanol. In 2019, we stained these specimens for bone using Alizarin Red.

We next replicated fossil data collection (Table 1) on these extant specimens. Standard length as well as pelvic-spine length on each side were measured with calipers. We used a dissection microscope to count dorsal spines, pelvic spines, dorsal-fin rays, and anal-fin rays. Right and left-side pelvic girdle lengths and ectocoracoid lengths were measured from ventral photographs taken using a Canon EOS Rebel T7 with a Tamron 16-300 mm MACRO lens mounted on a leveled Kaiser RS1 copy stand. Specimens were held in place for ventral photographs using a small tabletop vise with an attached scale bar. Lateral X-rays were used to measure dorsal spine length, number of pterygiophores anterior to the pterygiophore holding the third spine, length of the pterygiophore just anterior to the third spine, cleithrum length, and pre-maxilla ascending branch length. We also counted vertebrae from the X-rays: abdominal vertebrae were counted anterior to the first vertebra with a haemal spine contacting an anal fin pterygiophore. Caudal vertebrae were posterior, including the first vertebra with the haemal spine contacting the anal fin pterygiophore (following @Aguirre2014. X-rays were taken with an AXR Hot Shot X-ray Machine (Associated X-ray Corporation) at the Field Museum of Natural History. Specimens were exposed at 35kV and 4mA. Small fish were exposed for 7s, medium fish for 8s, and large fish for 10s. We developed the film and scanned individual images of each fish using the B&W Negatives setting on an Epson Perfection 4990 Photo flatbed at 2400 dpi. Measurements from photographs and X-rays were taken with FIJI (Schindelin et al. 2012) and its plugin ObjectJ (https://sils.fnwi.uva.nl/bcb/objectj/). All photographs, X-rays, and ObjectJ files have been uploaded to Morphosource.org (accession # TBD).
We dissected individuals from the generalist populations to determine sex from the gonads. Individuals from the species-pair lakes were sexed by Schluter and his lab, using a genotyping protocol (confirm and cite).

The extant data used here consists of a total of `r sum(!is.na(stick$gender))` specimens all with known sex.  Of these, there are `r sum(stick$gender == "Female",na.rm = TRUE)` and `r sum(stick$gender == "Male",na.rm = TRUE)` female and male specimens, respectively.

## Outlier analysis. 
To check for outliers, we calculated within-group means and standard deviations for each trait separately for K series fossil specimens (pooled across samples) and for extant specimens (pooled across lakes). We noted trait values greater than 3.5 standard deviations from the mean as potential outliers. We checked whether these potential outliers were a result of data entry and collection error and corrected them if they were. We turned the remaining outlier trait values to NAs. (confirm)  (Wait, what?  Why?  Anything that was 3.5 SD above the mean was recorded as missing?  How do you justify this?  )

Missing data imputation, including fossil sex

Quantification and evolution of sexual dimorphism

What covariates do we have in the data: length, what else,

>>>>

Ecological release theory suggests that a population’s niche should change when important species interactions (e.g., resource competition) are relaxed or removed (reviewed in @Herrmannetal2021). Removal is posited to create ecological opportunities—i.e., aspects of the niche become newly accessible, and the focal population shifts and/or expands its resource use to the new opportunity (@ParentandCrespi2009; @Herrmannetal2021). Ecological release may then be followed by adaptive morphological evolution (@ParentandCrespi2009; @Herrmannetal2021). 
For example, a population undergoing ecological release can experience disruptive selection on males and females stemming from intraspecific competition over newly accessible resources (@BolnickandDoebeli2003; @BolnickandLau2008; @Cooperetal2011). This is predicted to result in intersexual divergence in habitat use and associated phenotypes (@Schoener1968; @Shine1989; @BolnickandDoebeli2003; @Butleretal2007; @BolnickandLau2008; @Cooperetal2011; but see @Stuartetal2021; @Blain2022). Such competition-driven “intraspecific character displacement” between the sexes is therefore one explanation for the evolution of sexual dimorphism (@PfennigandPfennig2012; @DeLisleandRowe2015; @DeLisleandRowe2017, @De Lisleetal2018).

Much of the theory and empirical data for character displacement between the sexes is based on release from resource competition specifically (e.g., @BolnickandDoebeli2003; @Cooperetal2011; @PfennigandPfennig2012; @DeLisleandRowe2015; @DeLisleetal2018). However, release from predation should also result in ecological release followed by the evolution of increased sexual dimorphism because the absence of predators should generate ecological opportunity (@ReimchenandNosil2004; @ParentandCrespi2009; @Herrmannetal2021).

Here, we tested the prediction that a release from predation should result in the evolution of increased sexual dimorphism using a well-preserved, finely-resolved sequence of a fossil threespine stickleback fish (Gasterosteus doryssus). The sequence in this depositional environment is comprised of two lineages (@Bell2009; @Cerasonietal2024). Lineage I was a low-armor form with zero to one dorsal spines and highly reduced pelvises, on average. It lasted for at least 93,000 years before it was replaced suddenly by a second lineage (@Belletal1985). Lineage II appeared in the depositional environment fully armored, with complete pelvic girdles, two pelvic spines, and three dorsal spines, on average (@Belletal2006; @Stuartetal2020). It then immediately began evolving adaptive reduction in its armor traits (@Huntetal2008; @Stuartetal2020) until it reached the same low-armor state previously held by lineage I. 
The observation of 93,000 years of low-armor stasis in lineage I (@Belletal1985) and the observation of rapid evolution of armor loss by lineage II, both suggest that this depositional environment lacked piscivorous fish (e.g., trout and other salmonids known to prey on modern threespine stickleback). Armor maintenance in extant threespine stickleback (Gasterosteus aculeatus) correlates strongly the presence of vertebrate piscivores; populations with less predation pressure typically have less armor (@Reimchen1994, @ReimchenandNosil2004; @Belletal1993; @Roestietal2023). Consistently, only three fossil trout have been recorded from this paleolake basin, sampled from the same rock that has revealed >15,000 threespine stickleback fossils as well as occasional killifish (Fundulus nevadensis) (@Bell2009; @Cerasonietal2024). Finally, tooth wear data from the lineage II sequence suggest that individuals in this lineage began eating more planktonic prey over time, expanding toward an open-water niche from the benthic, bottom-feeding niche they started with (Purnell et al. 2007). That this expansion into open water by lineage II coincided with armor reduction is further indicative of a limnetic system with fewer salmonid piscivores (@SchluterandMcPhail1992; @VamosiandSchluter2004; @Roestietal2023). In sum, we infer that lineage II likely started in a nearby paleolake basin that had predators since it was armored when it arrived (@Bell2009; @Cerasonietal2024). Lineage II then experienced release from predators and dietary expansion in the focal paleolake basin, generating a scenario consistent with the starting conditions for models predicting the evolution of increased sexual dimorphism following ecological release.

A major challenge in the study of fossilized sexual dimorphism is assigning sex to individual specimens in the first place (@HoneandMallon2017; @Mallon2017; @Saittaetal2020). This typically cannot be done directly, except for taxa whose sexes are distinguished by the presence or absence of sex-specific characters that preserve well. Instead, paleobiologists often resort to statistical detection of sex and sexual dimorphism, including tests for normality and bimodality in trait distributions (e.g., @Mallon2017), mixture modeling (e.g., @Mallon2017), divergence in growth curves (e.g., @Saittaetal2020), and emphasis on effect size statistics rather than significance testing (e.g., @Saittaetal2020). However, dimorphic signal can be masked by noise introduced by factors both biological and artifactual, including extended growth during ontogeny, age-based bias in survivorship, small sample sizes, time averaging, and taphonomic bias (@Godfreyetal1993; @KoscinskiandPietraszewski2004; @HoneandMallon2017; reviewed in @Mallon2017; reviewed in @Saittaetal2020). Thus, the best approach to sex classification is likely one of total evidence (@Saittaetal2020), including comparison to closely-related, extant species of known sex (e.g., @HoneandMallon2017; @Saittaetal2020).

For our study, we inferred sex in G. doryssus fossils by comparing multivariate morphological trait data from lineage II samples to the same multivariate trait set collected from multiple populations in the closely-related, extant Threespine Stickleback species complex (Gasterosteus aculeatus). Crucially, we determined G. aculeatus sex directly via dissection and/or PCR genotyping. This enabled us to use Multiple Imputation by Chained Equations (MICE; @MICE) to build a predictive imputation model based on the multivariate morphology of G. aculeatus individuals of known sex. We applied this model to the fossil data to impute individual sex based on morphology, treating fossil sex as a missing variable. With sex assigned to fossil specimens (and accounting for uncertainty in those assignments), we tracked the multivariate evolution of sexual dimorphism over ~16,000 years of lineage II. We hypothesized that sexual dimorphism should increase following ecological release from predators.

@@

## Fossil Specimen Data
We used Gasterosteus doryssus data that were previously reported by @Stuartet al2020, @Voje etal2022, and @Siddiquietal2024. Briefly, the data were collected from fossil Series K from Quarry D (@Cerasonietal2024), dug from an open pit diatomite mine at 9.526° N, 119.094° W, near Fernley, Nevada, USA. Series K consisted of 18 samples taken at ~1000-year intervals, and mean sample times span ~16,363 years. Fish from series K were measured for 16 ecomorphological traits related to armor, swimming, and feeding (Table 1). Series K started at a previously documented horizon when a low armored lineage of stickleback with zero to one dorsal spines, zero pelvic spines, and highly reduced pelvises (i.e., lineage I described above) was completely replaced by a high armored lineage of stickleback with three dorsal spines, two pelvic spines, and complete pelvises (i.e., lineage II described above) (@Belletal2006; @Bell2009; @Stuartetal2020). Lineage II subsequently evolved reduction in armor, body size, and traits related to swimming and feeding (@Belletal2006; @Stuartetal2020; @Siddiquietal2024). The tempo and mode of armor reduction during this sequence suggests adaptive evolution by natural selection (@Huntetal2008), and we focus on the multivariate evolution of sexual dimorphism by this second lineage.

The lineage II fossil data consist of `r sum(is.na(stick$gender))` specimens of unknown sex sampled across 18 time periods spaced  ~1000 years apart. Figure 1 shows the sample size for each of the 18 samples.  There are at least 22 specimens in each sample with a high of 67 specimens in sample 7.

```{r fig.cap = "Fossil sample size for each time period"}
ggplot(aes(x = time, y = count), data = fortab) + geom_col() + xlab("Sample")  + scale_x_continuous(breaks = 1:18, labels = 1:18)
```

@@

## Extant Specimen data
To span the gamut of stickleback diversity for our predictive imputation model, we sampled modern stickleback from lakes containing generalist stickleback populations (@Hendryetal2009; @Bolnick2011) and from lakes containing benthic-limnetic species pairs (@Baumgartneretal1988; @SchluterandMcPhail1992). The generalist populations used here were the remainder of collections made by YES in 2013 and were previously described in @Stuartetal2017 (Table S1). These samples were fixed in formalin, then stained for bone with Alizarin Red in 2013. Benthic and limnetic specimens were kindly loaned by D. Schluter and his lab at University of British Columbia. They collected benthic and limnetic individuals from Enos Lake in 1988 and from Emily Lake, Little Quarry Lake, Paxton Lake, and Priest Lake in 2018 (Table S1). The Enos specimens had been fixed whole in formalin and stored in 40% isopropanol. The specimens from the other lakes were initially preserved whole in 95% ethanol in the field before being gradually transferred to water then formalin in the lab and ultimately stored in 40% isopropanol. In 2019, we stained these specimens for bone using Alizarin Red.
We next replicated fossil data collection (Table 1) on these extant specimens. Standard length as well as pelvic-spine length on each side were measured with calipers. We used a dissection microscope to count dorsal spines, pelvic spines, dorsal-fin rays, and anal-fin rays. Right and left-side pelvic girdle lengths and ectocoracoid lengths were measured from ventral photographs taken using a Canon EOS Rebel T7 with a Tamron 16-300 mm MACRO lens mounted on a leveled Kaiser RS1 copy stand. Specimens were held in place for ventral photographs using a small tabletop vise with an attached scale bar. Lateral X-rays were used to measure dorsal spine length, number of pterygiophores anterior to the pterygiophore holding the third spine, length of the pterygiophore just anterior to the third spine, cleithrum length, and pre-maxilla ascending branch length. We also counted vertebrae from the X-rays: abdominal vertebrae were counted anterior to the first vertebra with a haemal spine contacting an anal fin pterygiophore. Caudal vertebrae were posterior, including the first vertebra with the haemal spine contacting the anal fin pterygiophore (following @Aguirreetal2014). X-rays were taken with an AXR Hot Shot X-ray Machine (Associated X-ray Corporation) at the Field Museum of Natural History. Specimens were exposed at 35kV and 4mA. Small fish were exposed for 7s, medium fish for 8s, and large fish for 10s. We developed the film and scanned individual images of each fish using the B&W Negatives setting on an Epson Perfection 4990 Photo flatbed at 2400 dpi. Measurements from photographs and X-rays were taken with FIJI (@Schindelinetal2012) and its plugin ObjectJ (https://sils.fnwi.uva.nl/bcb/objectj/).

We dissected individuals from the generalist populations (Table S1) to determine sex from the gonads. Individuals from the species-pair lakes (TableS1) were previously sexed by the Schluter lab, using either dissection or a genotyping protocol (Whom, personal communication).

The extant data used here consists of a total of `r sum(!is.na(stick$gender))` specimens all with known sex (Table S1).  Of these, there are `r sum(stick$gender == "Female",na.rm = TRUE)` and `r sum(stick$gender == "Male",na.rm = TRUE)` female and male specimens, respectively.

## Outlier analysis and size correction.

To check for outliers, we calculated within-group means and standard deviations for each trait separately for K series fossil specimens (pooled across samples) and for extant specimens (within generalist, benthic, or limnetic categories). We noted trait values greater than 3.0 standard deviations from the mean as potential outliers. We deemed 3.0 s.d. to be a reasonable threshold for detecting errors without excluding biologically relevant values. We checked whether these potential outliers were a result of data entry and collection error and corrected them if they were. We turned the remaining outlier trait values to NAs. We size-corrected continuous traits only, as they varied with size, unlike count traits that are fixed during early development. We regressed each trait on standard length using a mixed-model regression, pooling all specimens, following @Stuartetal2017. We appended the size corrected data to the uncorrected trait data frame and used all of our data to build the imputation model described next.

## Missing data imputation, including fossil sex
We used multiple imputation (@little2002statistical) to impute the sex of the fossils many times.  Imputations were performed using the MICE (@MICE) algorithm and implemented in R (@R2022language) with M = 100 imputations performed. After imputing the sex of the fossil data, each completed data set is then used to fit a modified Ornstein–Uhlenbeck (OU) (@OUProcess) model using a Bayesian framework to look for sexual dimorphism across a variety of stickleback phenotypes.
